GeoSPARQL 1.1: an almost decadal update to the most important geospatial LOD standard

### 2021-03-13 manuscript.tex
further shortening

\author{
    Nicholas J. Car\inst{1}\orcidID{0000-0002-8742-7730} \and \\
    Timo Homburg\inst{2}\orcidID{0000-0002-9499-5840}\\
%    Simon J.D Cox\inst{3}\orcidID{0000-0002-3884-3420}
}

@@

\section{Motivation to update GeoSPARQL}\label{sec:motivation}
The OGC \& World Wide Web Consortium's (W3C) \textit{Spatial Data On The Web Working Group}
(SDWWG) established \textit{Spatial Data On The Web Best Practices}\cite{van_den_brink_best_2018} which noted 
shortcomings with then current spatial data standards: ``A best practice for returning geometries in a 
specific requested CRS has not yet emerged''. The group also informally captured specific suggested updates for 
GeoSPARQL\footnote{\url{https://www.w3.org/2015/spatial/wiki/Further_development_of_GeoSPARQL}}, however no updates 
to GeoSPARQL were then made.

Recently, 2019, the OGC reconstituted a \textit{GeoSPARQL Standards Working Group} (SWG) to update GeoSPARQL. The general 
motivation for work within the area of GeoSPARQL, that of \textit{Semantic Web} spatial data, and a series of
fault fixes and proposed extensions to GeoSPARQL 1.0 are captured in an OGC White Paper \cite{geosparqlwhitepaper}. Some,
but not all, of the SDWWG's ideas have been taken up by the SDW, for example the \textit{Best Practices}~\cite{van_den_brink_best_2018}
aspiration that ``A possible way forward is an update for the GeoSPARQL spatial ontology. This would provide an 
agreed spatial ontology, i.e., a bridge or common ground between geographical and non-geographical spatial data...''.
This point has been considered by the SWG and included in future releases' scope.

@@

The SWG's charter - its final scope of work - is also published by the OGC \cite{abhayaratna2020ogc} and this guides 
the SWG's activities. Specific actions of the SWG and their staging are explained through the use of a publicly-available 
online task tracking system within the SWG's working online code repository\footnote{\url{https://github.com/opengeospatial/ogc-geosparql/projects/1}}.

@@

\begin{itemize}
    \item[$\ast$] a revision of the ``upper  ontology'' structuring of GeoSPARQL
    \begin{itemize}
        \item[$-$] better/differently defining how GeoSPARQL's \texttt{Feature} and other classes relate to one another and to other fundamental concepts in the ontology
    \end{itemize} 
    \item[$\ast$] alignments to other ontologies, perhaps \textit{W3C Time Ontology in OWL}\cite{simon_cox_time_2017}
    \item[$\ast$] catering for very different SRSes, such as Discrete Global Grid Systems
\end{itemize}

@@

\subsection{Profile Declaration}\label{sec:profiledec}
One of the first actions undertaken by the SWG was to link the GeoSPARQL 1.0 elements through a \textit{profile} 
declaration, where a profile is a special type of \textit{specification}, as defined by \textit{The Profiles Vocabulary}\cite{atkinson_profiles_2020}. 
The specific motivation for this was the SWG's recognition that GeoSPARQL 1.0 consisted of multiple parts, not all
of which were easy to discover. As a result, some GeoSPARQL users were unaware of some of the resources and some
resources were accidentally duplicated or partly re-implemented. Profile declarations of this sort are anticipated, by the OGC, 
as being the \textit{best practice} way for it to deliver multi-part standards.

The profile declaration for GeoSPARQL 1.0 will be published by the OGC as a stand-alone resource sometime in early 2021 along with some 
updated GeoSPARQL 1.0 resources. The profile declaration for GeoSPARQL 1.1 will be published at the same time as all, or most, of the 
1.1 releases' updated resources, currently expected in mid-2021. The 1.1 releases' resources are:

\begin{enumerate}
    \item a profile declaration
    \item a specification document
    \item an RDF/OWL ontology document
    \item a Functions \& Rules vocabulary, derived from the ontology
    \item a Simple Features feature types vocabulary
    \item a set of RIF\cite{kifer2013rif} rules
    \item SHACL\cite{knublauch_shapes_2017} shapes for RDF data validation 
\end{enumerate}

Both the 1.0 and the 1.1 resources are currently available from the SWG's online code repository\footnote{\url{https://github.com/opengeospatial/ogc-geosparql}}.

@@

\begin{enumerate}
    \item \textbf{GeoJSON} (Geo- JavaScript Object Notation)\cite{butler2016geojson}
    \item \textbf{KML} (Keyhole Markup Language)\cite{nolan2014keyhole} 
    \item \textbf{DGGS} (Discrete Global Grid System)\cite{sahr1998discrete}
\end{enumerate}

@@

\begin{enumerate}
    \item No client-side library is needed to create an aggregated geometry result
    \item Fewer and more appropriate results can be returned using a spatial aggregate function (for example, a union geometry result)
    \item Spatial aggregates from different SPARQL endpoints can easily be calculated using federated queries 
\end{enumerate}

In addition to the \href{http://www.opengis.net/def/function/geosparql/union}{\emph{geof:union}}, \href{http://www.opengis.net/def/function/geosparql/envelope}{\emph{geof:envelope}} and \href{http://www.opengis.net/def/function/geosparql/convexHull}{\emph{geof:convexHull}} functions already defined in GeoSPARQL 1.0, 
for use within SPARQL \texttt{FILTER} operations, GeoSPARQL 1.1 defines \emph{geosf:Union}, \emph{geosf:Envelope} and \emph{geosf:ConvexHull}
as well as \emph{geosf:BoundingCircle}, \emph{geosf:Centroid}, \emph{geosf:ConcatLines} - concatenating a set of overlapping linestrings 
that overlap - and \emph{geosf:ConcaveHull} that can be used to return aggregated results. Listing~\ref{lst:agfunc} gives an example 
of one of these new functions in use.
% \todo[inline]{Timo: I will create a pull request in the GeoSPARQL repository to clear up the issues concerning the spatial aggregate functions. Then I will update the namespaces here. I would find it nice to also include the functions we have discussed yesterday: transformation of SRS and in between literal formats. Should we mention it here as future work or should we make a draft pull request for that so that people have a reference? I think the latter}
\small
\begin{lstlisting}[caption=Aggregation Function example SPARQL query,label=lst:agfunc,language=sql,frame=single,basicstyle=\ttfamily,showstringspaces=false]
# returns the circle geometry bounding all the geometries 
# of Feature <x>
SELECT (geosaf:BoundingCircle(?geo) AS ?circ)
WHERE {
    <x> geo:hasGeometry/geo:asWKT ?geo .
}
\end{lstlisting}
\normalsize

@@

These additions address a several requests from the SDWWG \& SWG but, given the introduction of a dedicated \textit{measure} class, 
do more too: they open up GeoSPARQL to general patterns of measurement present in ontologies 
such as the W3C's \textit{SOSA}~\cite{haller_semantic_2017}.

The 1.1 release also adds the property \href{http://www.opengis.net/def/function/geosparql/inSRS}{\texttt{geo:inSRS}}, which allows declarations of a geometry's SRS, independent of serializations which paves the way 
for future definition of SRSes in RDF, as anticipated for GeoSPARQL 2.0.

>>>>

\author{
    Nicholas J. Car\inst{1}\orcidID{0000-0002-8742-7730} \and \\
    Timo Homburg\inst{2}\orcidID{0000-0002-9499-5840}
%    Simon J.D Cox\inst{3}\orcidID{0000-0002-3884-3420}
}

@@

\section{Motivation to update GeoSPARQL}\label{sec:motivation}
The OGC \& World Wide Web Consortium's (W3C) \textit{Spatial Data On The Web Working Group}
(SDWWG) established \textit{Spatial Data On The Web Best Practices}~\cite{van_den_brink_best_2018} which noted 
shortcomings with then current spatial data standards: ``A best practice for returning geometries in a 
specific requested CRS has not yet emerged''. The group also informally captured specific suggested updates for 
GeoSPARQL\footnote{\url{https://www.w3.org/2015/spatial/wiki/Further_development_of_GeoSPARQL}}, however no updates 
to GeoSPARQL were then made.

Recently, 2019, the OGC reconstituted a \textit{GeoSPARQL Standards Working Group} (SWG) to update GeoSPARQL. The general 
motivation for work within the area of GeoSPARQL, that of \textit{Semantic Web} spatial data, and a series of
fault fixes and proposed extensions to GeoSPARQL 1.0 are captured in an OGC White Paper~\cite{geosparqlwhitepaper}. Some,
but not all, of the SDWWG's ideas have been taken up by the SDW, for example the \textit{Best Practices}~\cite{van_den_brink_best_2018}
aspiration that ``A possible way forward is an update for the GeoSPARQL spatial ontology. This would provide an 
agreed spatial ontology, i.e., a bridge or common ground between geographical and non-geographical spatial data...''.
This point has been considered by the SWG and included in future releases' scope.

@@

The SWG's charter - its final scope of work - is also published by the OGC~\cite{abhayaratna2020ogc} and this guides 
the SWG's activities. Specific actions of the SWG and their staging are explained through the use of a publicly-available 
online task tracking system within the SWG's working online code repository\footnote{\url{https://github.com/opengeospatial/ogc-geosparql/projects/1}}.

@@

\begin{itemize}
    \item[$\ast$] a revision of the ``upper ontology'' structuring of GeoSPARQL
    \begin{itemize}
        \item[$-$] better/differently defining how GeoSPARQL's \texttt{Feature} and other classes relate to one another and to other fundamental concepts in the ontology
    \end{itemize} 
    \item[$\ast$] alignments to other ontologies, perhaps \textit{W3C Time Ontology in OWL}~\cite{simon_cox_time_2017}
    \item[$\ast$] catering for very different SRSes, such as Discrete Global Grid Systems
\end{itemize}

@@

\subsection{Profile Declaration}\label{sec:profiledec}
One of the first actions undertaken by the SWG was to link the GeoSPARQL 1.0 elements through a \textit{profile} 
declaration, where a profile is a special type of \textit{specification}, as defined by \textit{The Profiles Vocabulary}~\cite{atkinson_profiles_2020}. 
The specific motivation for this was the SWG's recognition that GeoSPARQL 1.0 consisted of multiple parts, not all
of which were easy to discover. As a result, some GeoSPARQL users were unaware of some of the resources and some
resources were accidentally duplicated or partly re-implemented. Profile declarations of this sort are anticipated, by the OGC, 
as being the \textit{best practice} way for it to deliver multi-part standards.

The profile declaration for GeoSPARQL 1.0 will be published by the OGC as a stand-alone resource sometime in early 2021 along with some 
updated GeoSPARQL 1.0 resources. The profile declaration for GeoSPARQL 1.1 will be published at the same time as the 
1.1 releases, currently expected in mid-2021. All 1.0 and 1.1 relase resources are currently available in draft form 
in the SWG's online code repository\footnote{\url{https://github.com/opengeospatial/ogc-geosparql}}. The 1.1 releases' resources are:

\begin{enumerate}
    \item a profile declaration
    \item a specification document
    \item an RDF/OWL ontology document
    \item a Functions \& Rules vocabulary, derived from the ontology
    \item a Simple Features feature types vocabulary
    \item a set of RIF~\cite{kifer2013rif} rules
    \item SHACL~\cite{knublauch_shapes_2017} shapes for RDF data validation 
\end{enumerate}

@@

\begin{enumerate}
    \item \textbf{GeoJSON} (Geo- JavaScript Object Notation)~\cite{butler2016geojson}
    \item \textbf{KML} (Keyhole Markup Language)~\cite{nolan2014keyhole} 
    \item \textbf{DGGS} (Discrete Global Grid System)~\cite{sahr1998discrete}
\end{enumerate}

@@

\begin{enumerate}
    \item No client-side library is needed to create an aggregated geometry result
    \item Fewer/more appropriate results returned, for example a union result
    \item Federated SPARQL queries can aggregate results from multiple endpoints
\end{enumerate}

In addition to \href{http://www.opengis.net/def/function/geosparql/union}{\emph{geof:union}}, \href{http://www.opengis.net/def/function/geosparql/envelope}{\emph{geof:envelope}} and \href{http://www.opengis.net/def/function/geosparql/convexHull}{\emph{geof:convexHull}} defined in GeoSPARQL 1.0 
for use within SPARQL \texttt{FILTER} operations, 1.1 defines \emph{geosf:Union}, \emph{geosf:Envelope} and \emph{geosf:ConvexHull}
as well as \emph{geosf:BoundingCircle}, \emph{geosf:Centroid}, \emph{geosf:ConcatLines} - concatenating a set of overlapping linestrings 
that overlap - and \emph{geosf:ConcaveHull} that can return aggregated results. Listing~\ref{lst:agfunc} gives an example 
of one of these new functions in use.
% \todo[inline]{Timo: I will create a pull request in the GeoSPARQL repository to clear up the issues concerning the spatial aggregate functions. Then I will update the namespaces here. I would find it nice to also include the functions we have discussed yesterday: transformation of SRS and in between literal formats. Should we mention it here as future work or should we make a draft pull request for that so that people have a reference? I think the latter}
\small
\begin{lstlisting}[caption=Aggregation Function example SPARQL query,label=lst:agfunc,language=sql,frame=single,basicstyle=\ttfamily,showstringspaces=false]
# returns the circle geometry bounding all the geometries 
# of Feature <x>
SELECT (geosaf:BoundingCircle(?geo) AS ?circ)
WHERE {
    <x> geo:hasGeometry/geo:asWKT ?geo .
}
\end{lstlisting}
\normalsize

@@

These additions address requests from the SDWWG \& SWG but
do more too: they open up GeoSPARQL to general patterns of measurement present in ontologies 
such as the W3C's \textit{SOSA}~\cite{haller_semantic_2017}. Similarly, 
the 1.1 release addition of property \href{http://www.opengis.net/def/function/geosparql/inSRS}{\texttt{geo:inSRS}}, allows declarations of a geometry's SRS, independent of serializations and paves the way 
for future definition of SRSes in RDF, anticipated for GeoSPARQL 2.0.
